---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Git and GitHub for Collaboration

While version control with Git enhances reproducibility, GitHub takes it to the next level by facilitating collaboration. In this workshop, participants will practice collaborating on GitHub, including pushing and pulling changes to a shared repository. We will also explore some of GitHub’s additional functionality (e.g. using Issues) that can support scientific teams.

### Learning objectives 

The target audience for this workshop is any researcher looking to improve their use of GitHub for sharing and collaborating on research projects. Basic familiarity with Git, including using the command line and/or a Git client, is needed. However, we will cover the basics of syncing Git and GitHub. 

During the workshop, participants will

- Practice pushing and pulling changes to a remote repository
- Set up and make changes in a shared and collaborative repository
- Consider different kinds of GitHub workflows to facilitate collaboration

### Sources

The [GitHub documentation][] is a great resource.

A useful [comparison of workflows][] is here - note this is from BitBucket, which is an alternative Git hosting service to GitHub, but the ideas transfer.  

[Happy Git with R][] - gives a ton of guidance and trouble-shooting, useful for non-R users as well.

[Section 3][] of the Intro Earth Data Science textbook has detailed instructions for using Git and GitHub for scientific collaboration. 

[Section 3]: https://www.earthdatascience.org/courses/intro-to-earth-data-science/git-github/
[Happy Git with R]: https://happygitwithr.com
[GitHub documentation]: https://docs.github.com/en
[comparison of workflows]:https://www.atlassian.com/git/tutorials/comparing-workflows

### Slides

Slides presented during the workshop will be available [here](https://github.com/lillian-aoki/workshop_website/blob/master/slides/04_github_collaboration.pdf)


## Getting started

The first building block of using GitHub for collaboration is making sure that Git, GitHub, and your Git client can all connect and interact. Note, you don't have to use a Git client, but it can make certain elements easier as you start working with GitHub. 

We will run through a series of steps to create a remote repository, clone (download) to your local machine, and make sure Git is tracking changes made locally and in the remote repository. 

These steps will also involve setting up access to GitHub using either HTTPS or SSH keys. This is slightly fiddly if you've never done it before.  

### Set up GitHub access

There are two options for setting up credentials for GitHub to talk to your local machine: SSH or HTTPS. Using HTTPS means you need to create a personal access token, which operates much like a password. Using SSH requires generating a  
pair of keys (two long sets of character strings).

If you've never used GitHub before, HTTPS and a personal access token is probably slightly easier to get started. 

Follow the [instructions on this page][] to generate the personal access token (PAT).  

You can name the PAT something like 'my laptop' or 'research workshop' - a short name that gives you some info about how you are using it. When you are asked to select the scope, select "repo", "user" and "workflow". 

[instructions on this page]:https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/creating-a-personal-access-token

<p class="text-success">
Note, when you generate the PAT, keep your browser window open. You won't be able to access the PAT again once you've navigated away. For the purposes of this workshop, copy the PAT to a text file so that you'll be able to copy and paste it as needed.  
</p>

If you prefer to use SSH, you can follow the [instructions provided here][]. Note that you have an option to add another layer of security to SSH by including a passphrase. for the workshop, I would recommend *not* including a passphrase. You can add it later if you feel so inclined. 

[instructions provided here]: https://docs.github.com/en/authentication/connecting-to-github-with-ssh

### Create a remote repository

The next step is to create a new repository. Return to you GitHub account in your browser and click the green `New repo` button. Make the following selections:  

- No template
- Repository name: use something short like 'my_repo' or 'test_repo'
- Add a line to the description (this will appear in your ReadMe file)
- Public
- Add a ReadMe file

Click `Create repository`. GitHub will immediately create that new repo, with the text from your description displayed in the ReadMe. 

Next, you want to clone this remote repo to your local machine. Click the green `Code` button in the upper right of screen. You'll need to select HTTPS or SSH, depending on the credentials you are using. Next, click the button with two squares  
to copy the URL.

### Connecting your local repo to a remote

Open the terminal, and navigate to a logical place in your filesystem for this test repo. For example, you could use folder `temp` in Desktop or Documents.  

Clone the remote repo to your computer using the following syntax:

```{r, eval=FALSE, purl=FALSE}
$ git clone <paste the URL you copied>
```

The pasted URL should include your GitHub username and the name of the repo, e.g.  

`https://github.com/YOUR-USERNAME/YOUR-REPOSITORY.git`

If you are using SSH instead, the URL will use the format  

`git@github.com:YOUR-USERNAME/YOUR-REPOSITORY.git`

You'll see some output stating the result of cloning the directory. 

Now, make this new repo yoru working directory by navigating into it, then list the files, and view the head of the Readme file

```{r, eval=FALSE, purl=FALSE}
$ cd test_repo  
$ ls  
$ head README.md  
```

The output from these commands should be familiar from the basics of working in the shell. Next, show some information about the connection to Github by typing:  

```{r, eval=FALSE, purl=FALSE}
$ git remote show origin
```

You should see output similar to the following (note the URLs should match the URL you copied before using the `clone` command)

```{r, eval=FALSE, purl=FALSE}
  `* remote origin  
  Fetch URL: git@github.com:lillian-aoki/new-test.git  
  Push  URL: git@github.com:lillian-aoki/new-test.git    
  HEAD branch: main  
  Remote branch:  
    main tracked  
  Local branch configured for 'git pull':  
    main merges with remote main  
  Local ref configured for 'git push':  
    main pushes to main (up to date)  
```

### Pushing and pulling changes

Once you've connected your local machine to the remote repo on GitHub, you can start pushing and pulling changes between the local and remote branches.  

From the command line, we can add a new line to the Readme:

```{r, eval=FALSE, purl=FALSE}
$ echo "A line I wrote on my local computer" >> README.md
```

Check the `git status`, then use `git add` to stage the changes and `git commit` to commit the changes. Remember to use the `-m` option with `commit` and to write a succinct commit message, like `Add a change from my local computer`

Once the changes have been committed, check the `status` again. You should see a line stating that your branch is ahead of 'origin/main' by 1 commit. This means the local copy of the repository has changes that have not been propagated to the remote copy on Github. Propagate changes using the `push` command

```{r, eval=FALSE, purl=FALSE}
$ git push 
```
At this point, you will likely be asked for your username and password, if you are using HTTPS. Enter your GitHub username and the PAT (copy and paste from your text file). If you are using SSH keys, you will not need to enter a password unless you set up extra security with a passphrase for SSH.

Now when you check `git status` it should state that your branch is up to date with origin/main, i.e. the local and rmeote branches are synced. 

Go to your browser and refresh the repository page. You should see the edited text appear in your ReadMe. 

We can also propagate changes from the remote copy to the local copy of the repo. Click the pencil icon in the corner of your ReadMe file to edit the file in your browser. Add another line of text, such as 'A new line from the remote repo'.  

Scroll to the bottom of the screen, and you'll see an option to add a commit message. You can leave the default message, or edit it, then click `Commit changes`. 

Return to the homepage of your repository, and you'll see new information with the commit identifier and message indicating which files were edited in this commit.  

If you click on the number of commits in the upper right, you'll see a change log, showing the commits you've made via GitHub and on your local computer. This visual might be easier to understand than using `git log` in the command line! 

Now, we need to propagate the changes from this commit to the local copy. Return to the command line and use the `pull` command:

```{r, eval=FALSE, purl=FALSE}
$ git pull  
```
This will download the changes recorded in the last commit you made locally. Use `cat` or another command to view the ReadMe file to verify that the changes have propagated.

Note that `git pull` is actually performing two actions, first downloading the changes to files from the remote repo, then integrating those changes into the local files. You can also perform these actions individually using `git fetch` to download changes and `git merge` to merge those changes with local files.  

As you are starting out getting used to `git`, using `git pull` is less likely to cause issues, but it can be useful to know that these actions can be separated. 

### Remote as back-up

Once you've connected your repo to the remote on GitHub, you can think of the version on GitHub as your backup. If something happens to your local files or computer, you can always clone the remote repo from GitHub. This means you'll want to push your local commits to GitHub regularly, so that the version on GitHub stays up to date. 

Generally, it's good practice to pull any changes from GitHub before pushing a new commit. Your standard workflow may be to make changes locally, saving changes to your files as you go. Once you've made a substantive change, commit that change locally. Then, when you want to propagate to GitHub, first pull any new changes from the remote, then push your recent commits. This can be especially important if multiple users are making changes locally and pushing to the same remote.  

### Using branches

Branches are an important feature of Git. Each branch is a series of commits, independent from the commits in the main branch. You can use branches to focus on developing a particular area of your project incrementally. Multiple users can also work in parallel branches without overwriting each other's work. 

However, it is *very* easy for branches to get messy. Multiple collaborators can develop branches in different ways that make it hard to merge those branches back into the main stream of the repo. When you create a branch, it's good to have a  
clear idea of what you want to accomplish with that branch, and when you will merge it back into the main branch. Short branches that are easily and frequently merged will make collaborating easier rather than harder. 

You can easily create a branch using the command `branch` and give it any name. For example, to create a new branch in your repo related to a specific issue in the code, use 

```{r, eval=FALSE, purl=FALSE}
$ git branch issue-1
```
Switch to the new branch using `checkout`:

```{r, eval=FALSE, purl=FALSE}
$ git checkout issue-1
```
You can also do these two steps in one line: 
```{r, eval=FALSE, purl=FALSE}
$ git checkout -b <branch name>  
```
In this new branch, you can develop code to address a specific issue or to advance part of your analysis. You should commit your work regularly on the branch as you would on the main branch. As a demonstration, let's add another line to the README file, although this would not actually address a coding issue in real life!

```{r, eval=FALSE, purl=FALSE}
$ echo "A new line added from the feature branch issue-1" >> README.md
```

We can then stage and commit this edit as usual:  

```{r, eval=FALSE, purl=FALSE}
$ git add README.md  
$ git commit -m "Add edit from new branch"  
```
When you are ready to merge the branch back into `main`, first switch to the main branch, and then merge the feature branch:  

```{r, eval=FALSE, purl=FALSE}
$ git checkout main  
$ git merge issue-1
```
Ideally, this will go smoothly. However, you might run into a merge conflict if the branches you want to merge contain edits to the same parts of the same files.  

### Merge conflicts

Merge conflicts are often small and easily resolved. You will get an error message if you have a merge conflict, stating that the automatic merge has failed. You can then use `status` to identify which files have conflicts. 

Open the file with a merge conflict, e.g. in your text editor. You'll see that the areas of conflict are highlighted with makers: `<<<<<<` indicates the beginning of the conflict, followed by content from the branch you are currently on (e.g. `main`). The marker `======` indicates the end of content from the `main` branch and the start of content from the feature branch (e.g. `issue-1` in this example). The end of the conflicting content is marked by `>>>>>`. 

Edit the content between the `<<<<<` and `>>>>>` markers to the version you want to keep. Make sure to delete the conflict makers and any content you do not want to keep. 

Then, make sure to stage and commit these changes using `git add` and `git commit`. This completes the merge.  

The following steps will show how to create an intentional merge conflict so that you are familiar with the syntax.  

First, make an edit too the first line of your README file on the main branch. Add the text 'New text for merge conflict'. You can do this with `nano`. Then use `git add` and `git commit` to stage and commit this change.  

Now switch to your feature branch using 
```{r, eval=FALSE, purl=FALSE}
$ git checkout issue-1
```

Edit the same file, the README file using nano. Delete the text you added before and write 'Different text to create conflict'. Stage and commit this change on the feature branch.  

Next, return to the main branch and merge the two branches:  
```{r, eval=FALSE, purl=FALSE}
$ git checkout issue-1 
$ git merge
```

You should receive a warning the merge failed.  

Use `git status` to check which file failed to merge, then open that file with `nano`. You'll see the markers showing where the conflict occurred. Edit the file to eliminate the conflict, save and close. 

Finally, stage and commit the changes you just made. This commit completes the merge you started earlier. If you run `git merge` after the commit, you'll see that the files are up to date.  

If you get into trouble during a merge, you can run `git merge --abort` to stop the merge and go back to the moment before you started merging.  

## Working with others

### What are your needs for collaboration?

When using GitHub for collaboration, it's important to consider what your needs are.  
Some examples are below:  

- A grad student creates a repo to share analysis results with their PI  
- A research group uses a shared repo to manage data collected annually in a long-term project, with different contributors over time  
- Authors create a GitHub repo to publish code as part of an open-access manuscript  
- A researcher develops a new package with tools for analysis or visualization and shares via GitHub
 
In these examples, the needs of the collaborators may be different, and therefore the workflow and set-up of a collaboration through GitHub may also differ. Some things to consider as you incorporate GitHub into your research workflow:  

- Who are your collaborators?  
- What do they need access to - data, code, analysis results?  
- What will your collaborators contribute - cleaning data, writing code, etc?  

As always, include your future self as a collaborator! 

### Fork-and-clone workflow

If you are working on a project with collaborators who will be contributing, the fork-and-clone workflow is a common approach. This workflow allows multiple users to contribute to a shared project, to keep track of each others changes, and to flag issues. 

For purposes of the workshop, we can practice the fork-and-clone workflow using the demo GitHub repo `practice-workflow`  

Navigate to the home page of the `practice-workflow` repo on Github.

In the upper right, click on `Fork` to create a fork of the repo. This will fully copy the contents of the repo to _your_ github account. 

Clone the forked copy from _your account_. Do so by clicking the green button, choosing HTTPS or SSH as appropriate to copy the URL. Then navigate to the command line, and in a reasonable directory on your computer, create the new repo using 

```{r, eval=FALSE, purl=FALSE}
$ git clone <URL>
```

At this point, you have a local copy of the repo on your machine. This repo is tracking the remote copy on your GitHub account. You can commit changes to your local repo and push those changes to your remote on GitHub, following the same procedures as before. In addition, because the remote on GitHub is a fork of the source repository, you can send changes that you make back to the source repository as pull requests. 

### Pull requests

A pull request submits changes to a shared project that you've made in your fork of the project repo back to the source repo (where the project ultimately lives). When you submit a pull request, whoever maintains the source repo (e.g. your  
collaborator), can review the changes and decide to incorporate them into the source repo. 

Make some local changes to your repo by following the instructions in the file `exercise.md`. You will make a new branch, then create a new file with some demo text. Commit this change to your local repo, then push the change to your remote repo.

Initiate a pull request by navigating to homepage of your remote repo, and clicking `Pull request` in the upper right. 

You need to identify the _base_ repository and the _head_ repository. The base is the repo that you want to update (the source repo you had forked earlier) and the head is the repo with content you want to use in the update (your forked copy). 
Then click `create new pull request`. You will need to add a title and description to your pull request - like a commit message, the tile should be short but informative.  

Finally click the green `Create pull request`. This will create the request and submit it to the source repo. 

Your colleague who maintains the source repo can then review the changes in the request before merging with the repo. You can also use the GitHub interface to leave comments to adjust changes before finalizing the changes and merging. 

### Using branches in fork-and-clone

An important point about the fork and clone workflow is that it is best practice to do your work in a feature branch, not the main branch. For example, in the exercise, you made a new branch named `exercise-YOURNAME`, then went through the the whole workflow (commit, push, pull request) from that branch. 

There are two reasons to work in a feature branch with fork and clone. First, it keeps your main branch clean so that you can pull any new changes from the upstream source repo. This is covered below. Second, it is slightly easier to manage pull requests that come from feature branches. 

However, using branches does increase the complexity of the workflow, which can create a challenge to incorprating GitHub into a team collaboration. 

### Getting upstream changes

A final element of the fork and clone workflow is that you will likely want to set up your fork to track changes that occur in the source repo. For example, if new data are added over time, or other collaborators contribute changes to the source repo, you want to keep your local copy up to date. 

The way to do this is to add the source repo as another remote to your local repo. 

In the shell, navigate into your local copy of the repo of interest. Check the status to make sure you are on the main branch and your working tree is clean (no changes since the last commit). 

List your current remotes: 
```{r, eval=FALSE, purl=FALSE}
$ git remote -v
```

You should see something like  
```{r, eval=FALSE, purl=FALSE}
origin <URL_for_your_repo> (fetch)  
origin <URL_for_your_repo> (push)
```
In a web browser, navigate to the *source* repo (the one you forked from before).  

Use the green 'Clone or download' button to copy the URL for the source repo.

Add this URL as a remote:
```{r, eval=FALSE, purl=FALSE}
$ git remote add upstream <URL_for_source_repo>
```
Now, check your remotes again using `git remote -v`

You should see output with new lines indicating the upstream remote
```{r, eval=FALSE, purl=FALSE}
origin <URL_for_your_repo> (fetch)  
origin <URL_for_your_repo> (push)  
upstream <URL_for_source_repo> (fetch)  
upstream <URL_for_source_repo> (push)
```
Now you can pull changes from the upstream remote into your local copy and keep everything up to date. 

### Other workflows

Fork and clone is a popular workflow for open source projects and is promoted by various teaching resources for scientists implementing Git. However, it can introduce complexity into your workflow, especially if everyone on your team is learning to use Git. It's important to balance the utility of GitHub with the time and resources (including cognitive load) required to implement these tools. 

I highly recommend the BitBucket tutorial [comparing workflows][] to compare some alternative structured workflows.

[comparing workflows]:https://www.atlassian.com/git/tutorials/comparing-workflows

I also recommend that researchers who are starting out with Git and GitHub focus on getting a basic workflow in place for themselves before building a collaborative project. There's huge utility in having a GitHub repo that serves as a backup,  allows you to share code, outputs, and visualizations easily, and imposes some deliberate organization on your project. Even if you do not have any collaborators working on the same repo, there's still a lot to gain from using GitHub. 

### Giving access

You may need to give collaborators access to a repository. The default setting for your repos is Public, which means anyone can view it. If you want specific collaborators to be able to commit directly to your repository, you can change the permissions. 

In your browser, navigate to your repository and go to Settings (in the menu bar for the repo). Click on `Manage access` and then click on the green button to add people. You will need their GitHub username or email to add them. 

You might also have a private repository, which is not visible to the public. You will need to add collaborators to the private repository in order for them to view the content. 

## Project Management in GitHub

Beyond the structure of version control, GitHub also provides some tools for project management. These can be fairly basic, such as using Issues (or the new Discussions) to faciliate communication about the project. There are also additional tools for project management, such as project boards (similar to other project management programs like Trello) and assigning milestones. These tools may become more useful for larger projects with many collaborators who are all using GitHub.

### Using Issues

Issues provide a way to flag elements of the project that need attention, and also provide an avenue for discussion with your collaborators. It can be helpful to use issues to discuss changes, then when you submit a pull request you can reference the issue that your changes address.

Practice using issues in the practice-workflow repo by going to `Issues` in the upper left of the menu bar. You'll see an open issue, essentially as a comment chain. 

Add a new comment to the open issue.

You can also use tags to organize issues, and integrate them with email for improved communication. GitHub has recently introduced Discussions, which functions similarly to Issues but are oriented toward more general discussion, rather than bug fixes. 
